---
title: "Fun with anonymous functions in R"
date: "2026-06-22"
categories:
  - Metaprogramming
  - R
description: A fun distraction to make an alternate syntax for anonymous functions in R.
engine: knitr
#doi: 10.59350/mdxxv-ke330
---

Julia has an elegant syntax for making anonymous functions: `(x, y) -> x^2 + x^2`. R's `\(x, y) x^2 + y^2` syntax isn't bad either, but I wanted to try and make something similar to Julia's syntax as an excuse to play around with R's metaprogramming facilities. It's quite possible that there's an R package out there that already does it---and I did look, briefly, without finding anything---but this is more of an excuse to do some metaprogramming than a serious need for a different way of writing out anonymous functions.

Let's start by making a new, minimal infix operator.

```{r}
`%->%` <- function(lhs, rhs) {
  res <- function() {}
  formals(res) <- as.pairlist(lhs)
  body(res) <- substitute(rhs)
  environment(res) <- parent.frame()
  res
}

Map(alist(x =, y = 1) %->% {x^2 + y^2}, 1:3)
```
Not bad. The curly braces make it look a bit more like Rust's syntax. But having `alist()` on the left-hand side seems ugly. What we can do instead is have something like `c(x, y = 1) %->% {x^2 + y^2}`.

```{r}
`%->%` <- function(lhs, rhs) {
  res <- function() {}
  # Extract the lhs args and set names.
  args <- as.list(substitute(lhs))[-1]
  unnamed <- if (is.null(names(args))) {
    seq_along(args)
  } else {
    which(!nzchar(names(args)))
  }
  names(args)[unnamed] <- lapply(
    args[unnamed],
    \(x) if (is.symbol(x)) as.character(x) else stop("invalid lhs")
  )
  args[unnamed] <- list(quote(expr =))
  # Same as before.
  formals(res) <- as.pairlist(args)
  body(res) <- substitute(rhs)
  environment(res) <- parent.frame()
  res
}

Map(c(x, y = 1) %->% {x^2 + y^2}, 1:3)
```

That seems about as close as we're going to get. Although this isn't particularly useful, it's neat to see how flexible R can be.
